Sequential rank agreement methods for comparison of lists

### 2015-07-23 latex/paper.Rnw
Updated manuscript

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[]{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[]{color}
\usepackage{subcaption}
\usepackage{natbib}
\usepackage{xspace}
\usepackage{calc}
\usepackage[noae]{Sweave}
\usepackage{authblk}
\usepackage{algorithm}
\usepackage{algpseudocode}
%
% New commands
%
\newcommand{\ie}{\emph{i.e.}\@\xspace}
\newcommand{\eg}{\emph{e.g.}\@\xspace}

@@

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and prominent task in applied statistics. Predictor
  variables are ranked according to magnitude of association with an
  outcome, prediction models rank subjects according to the
  personalized risk of an event, and genetic studies rank genes
  according to their difference in gene expression levels. This
  article constructs measures of the agreement of two or more ordered
  lists. We use the standard deviation of the ranks to define a
  measure of agreement that both provides an intuitive interpretation
  and can be applied to any number of lists even if some or all are
  incomplete or censored. The main idea is then a permutation based
  and graphically approach to assess the sequential changes of
  agreement as a function of the depth of the lists. This may help to
  indentify change points in the agreement of the lists. The
  usefulness of these tools are illustrated using gene rankings, and
  using data from two Danish ovarian cancer studies on where we assess
  the within and between agreement of different statistical classification
  methods.
\end{abstract}

Key words: sequential rank agreement, order statistic, gene rankings,
methods comparison, variable selection, permutation

@@

Several approaches exist to measure the ``distance'' between two
ranked lists. Of these, Kendall's $\tau$ \citep{kendall1948} and
Spearman’s footrule $\rho$ \citep{Spearman1910} are among the most
well-known. But these measures do not distinguish between agreement in
the top versus towards the bottom of the lists. \citet{Shieh1998}
proposed a weighted version of Kendall's $\tau$ where each pair of
rankings can be assigned different weights, and \citet{Yilmaz2008}
proposed the $\tau_{ap}$ which places higher emphasis on the top of
the lists.
%
Spearman’s footrule uses the ranks of the variables for calculation of
the distance and the use of ranks are also employed in the $M$ measure
of \citet{Bar-Ilan2006} where the reciprocal rank differences are used
to calculate the agreement measure.

@@

In this article we introduce sequential rank agreement for measuring
agreement among ranked lists. We provide graphical tools to analyse
the depth of agreement of a set of ranked lists. The general idea is
to define agreement based on the sequence of ranks from the first $d$
elements in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{Carstensen2010}. Our proposed approach allows to compare more
than two lists simultaneously, provides a dynamic measure of agreement
as a function of the depth in the lists, places high weight on the top
of the list, accommodates censored/incomplete lists of varying
lengths, and has a natural interpretation that directly relates to the
ranks. The graphical illustration of sequential rank agreement
potentially allows to infer a changepoint, i.e., a list depth where
there is a change in the agreement of the lists.

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$. An
ordered list is a permutation function, $R: \{X_1,\dots,X_P\}\to
\{1,\dots,P\}$, such that $R(X_p)$ is the rank of item $X_p$ in the
list. The inverse mapping $\pi=R^{-1}$ assigns to rank
$r\in\{1,\dots,P\}$ the item $\pi(r)$ found at that rank. The methods
described below work for a set of $L$ lists $R_1,\dots,R_L$,
$L\geq2$. We denote $\pi_l=R_l^{-1}$ for the corresponding inverse
mappings. Panels (a) and (b) of Table~\ref{tab:example} show a
schematic example of these mappings. Thus if $\pi_l(1)=X_{34}$ then
item $X_{34}$ is ranked first in list $l$, and similarly
$R_l(X_{34})=1$.

\begin{table}[tb]
  \caption{Example set of ranked lists. (a) shows the ranked list of items for each of three lists, (b) presents the ranks obtained by each item in each of the three lists and (c) shows the cumulative set of items up to a given depth in the three lists.}
\begin{center}
  \begin{subtable}{4cm}%
    \caption{}
      \begin{tabular}{cccc}
        \hline\hline % & \multicolumn{3}{c}{List} \\
        Rank & $\pi_1$ & $\pi_2$ & $\pi_3$ \\ \hline
        1 & A & A & B \\
        2 & B & C & A \\
        3 & C & D & E \\
        4 & D & B & C \\
        5 & E & E & D \\ \hline
    \end{tabular}
  \end{subtable}
% \hfill
\hspace{1em}
%%%
  \begin{subtable}{4cm}%
    \caption{}
    \begin{tabular}{cccc} \hline\hline
  % & \multicolumn{3}{c}{List} \\
    Item & $R_1$ & $R_2$ & $R_3$ \\ \hline
    A & 1 & 1 & 2 \\
    B & 2 & 4 & 1 \\
    C & 3 & 2 & 4 \\
    D & 4 & 3 & 5 \\
    E & 5 & 5 & 3 \\ \hline
  \end{tabular}
\end{subtable}
% \hfill
\hspace{1em}
%%%
\begin{subtable}{4cm}%
  \caption{}
\begin{tabular}{cc} \hline\hline
% Depth   &     \\
Depth  &  $S_d$ \\ \hline
1 & $\{$A, B$\}$\\
2 & $\{$A, B, C$\}$ \\
3 & $\{$A, B, C, D, E$\}$ \\
4 & $\{$A, B, C, D, E$\}$ \\
5 & $\{$A, B, C, D, E$\}$ \\ \hline
\end{tabular}
\end{subtable}
\end{center}
\label{tab:example}
\end{table}

The agreement of the lists regarding the rank given to an item $X_p$ can be
measured by
\begin{equation}
  A(X_p) = f(R_1(X_p), \ldots, R_L(X_p)),
\end{equation}
for a distance function $f$. Throughout this paper we will use the
sample standard error as our function $f$ and hence use
$$A(X_p) = \sqrt{\frac{\sum_{i=1}^L (R_i(X_p) - \bar{R}(X_p))^2}{L-1}},
$$
but other choices could be made (see the discussion). The sample
standard error has an interpretation as the average distance of the
individual rankings of the lists from the average ranking.

We now describe what is exemplified in Panel (c) of Table
\ref{tab:example} and how it can be used to define \emph{sequential
  rank agreement}. For an integer $1\le d\le P$ we define the unique
set of items found by merging the first $d$ elements of each of the
$L$ lists, i.e., the set of items ranked less than or equal to $d$ in
any of the lists:
\begin{equation}
S_d = \{\pi_l(r) ; r \leq d, l = 1, \ldots, L \}.
\end{equation}
The \emph{sequential rank agreement} is the pooled standard deviation
of the items found in the set $S_d$:
\begin{equation}
\textrm{SRA}(d)= \sqrt{\frac{\sum_{\{p \in
      S_d\}}(L-1)A(X_p)^2}{(L-1)|S_d|}}. \label{def:sra}
\end{equation}
Values of \textrm{SRA} close to zero suggests that the lists agree
while larger values suggests disagreement. If the ranked lists are
identical then the value of SRA will be zero for all depths $d$. The
sequential rank agreement can be interpreted as the average distance
of the individual rankings of the lists from the average ranking for
each of the items we have seen until depth $d$. Note that the terms
that are part of \eqref{def:sra} are not independent since each list
contains the ranks from 1 to $P$ exactly once. Hence we simply use the
pooled standard deviation as a measure of the average distance between
rankings.

@@

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to classify between acute lymphoblastic leukemia
(ALL) and acute myeloid leukemia (AML)
\citep{Golub1999}. Preprocessing of the gene expression data was done
as described in \citet{Dudoit2002} and the four different analysis
approaces were: marginal two-sample $t$ tests, marginal logistic
regression analyses, logistic regression eleastic net
\citep{friedman2010regularization}, and marginal maximum information
content correlations (MIC) \citep{Reshef2011}. For the first two
methods, the genes were ranked according to $p$ value, for logistic
regression the genes were ordered by size of the corresponding
coefficients (after standardization), and MIC was ordered by
correlation which resulted in the rankings seen in Table~\ref{tab1}.
The sequential rank agreement shown in Figure~\ref{fig:example1} shows
for each depth $d$ how many ranks apart do we on average expect the
genes found in the top $d$ part of the lists. It can be seen that
agreement is higher in the top of the lists (smaller values on the $y$
axis) correspond to \emph{better} than in the tails of the lists.

@@

If the lists would agree on all items then we would expect the
sequential rank agreement curve to be flat. And if there are changes
in the levels of rank agreement then this would suggest that there are
sets of items for which the lists are in high agreement and sets of
other items where the lists rank the items differently. In gene
association studies we generally expect the agreement among the lists
to be better towards the top of the lists and worse towards the end of
the lists. In this case the sequential rank agreement curve will start
at a low level and then increase until it levels off as in
Figure~\ref{fig:example1}.

@@

Incomplete lists are also a common occurrence. They arises for example
in case of missing data (items) or when some methods only rank a
subset of the items. For example, penalized regression based on the
Lasso provides a sparse set of predictors that have non-zero
coefficients. There is no obvious ordering of the set of predictors
whose coefficient has been shrunk to zero. Incomplete lists also occur
when the analyst decides to censor all lists at depth $d$ such that
the measurement of agreement is restricted to the first $d$ elements
of each list.

Sequential rank agreement can be generalized to incomplete/censored
lists in the following way. Let $\Lambda_P$ be a set of $P$ items and
$\Lambda_l\subset \Lambda_P$ the subset of $d_l$ items that occur in
list $l$.  The case where all lists are censored at depth $d$
corresponds to $d_1=\cdots=d_L=d$. For incomplete/censored lists the
rank function becomes
\begin{equation}
  \tilde R_l(X_p) = \left\{\begin{array}{cl} \{\pi_l^{-1}(p)\} & \text{ for } p\in \Lambda_l \\
      \{d_l+1,\dots,P\} & \text{ for } p \not\in \Lambda_l\end{array}\right.
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

@@

Like for fully observed lists we generally expect the sequential rank
agreement to start low and then increase unless the lists are
completely unrelated (in which case the sequential rank agreement will
be constant at a high level) or if the lists mostly agree on the
ranking (in which case the sequential rank agreement will also be
constant but at a low level). For censored lists we also expect a
change point around the depth where the lists are censored. This is an
artefact stemming from the fact that we assume that the remainder of
the lists can be replaced by a simple permutation of the missing
items.

@@

The first reference hypothesis is
\begin{eqnarray*}
H_0  & : &  \text{The list rankings correspond to complete randomly}\\
       &  & \text{permuted lists}.
\end{eqnarray*}
This does not only assume that there is no information data on which
in the rankings are based but also that the methods are completely
independent.

@@

The overlap among $L$ lists observed until depth $d$ is defined as the
proportion of items that are found in all $L$ lists compared to the
possible number of items found in all lists observed until depth $d$:
\begin{equation}
  O(d) = \frac{| \cap_{l=1}^L \{\pi_l(r); r\leq d \} |}{d}. \label{def:overlap}
\end{equation}
Here $|A|$ is the cardinality of the set $A$.  Typically the overlap
is only considered for pairwise comparisons but formula
\eqref{def:overlap} can be applied even when $L>2$
\citep{Bar-Ilan2006,Boulesteix2009}.

@@

The MALOVA study is a multidisciplinary Danish study on ovarian cancer
\cite{Hogdall:2004:Cancer:15160342} where all Danish women diagnosed
with an ovarian tumor and referred for surgery from the participating
departments of gynecology were enrolled continuously from December
1994 to May 1999.
% @Andreas: hvorfor random subsample?
We use a random subsample of $119$ patients with a total of $58$
patients with malignant ovarian cancers as cases and $61$ patients
with benign ovarian tumors as controls. The DACOVA study is another
multidisciplinary Danish study on ovarian cancer which included about
$66\%$ of the female population of Denmark
\cite{bertelsen1991protocol}. It was aimed to continuously enroll all
patients that were referred to surgery of an ovarian tumor clinically
suspected to be cancer during the period 1984 to 1990.
% @Andreas: hvorfor random subsample?
For the purpose of illustration use a random subsample of $113$
patients with a total of $54$ malignant ovarian cancers and $59$
benign ovarian tumors/gynecologic disorders.

Proteomic analysis was performed by MALDI-TOF mass spectrometry with a
Bruker Ultraflex in positive, linear ion mode in the mass-to-charge
range $800$ to $20000$ Dalton by an average of $500$ laser
shots. Calibration of the instrument was performed using a protein
standard mixture. Each spectrum consists of $49642$ samples over the
mass-to-charge range which we down-sample on an equidistant grid of
5000 points by linear interpolation. We further preprocess each
spectrum by first removing the slow-varying baseline intensities with
the SNIP algorithm \cite{ryan1988snip} followed by a normalization
with respect to the total ion counts. Finally, we standardize the 5000
predictors to have column-wise zero mean and unit variance.

We use these two data sets to illustrate how the sequential rank
agreement can be applied in two different scenarios. In the first
scenario we assess the agreement of different statistical
classification methods in how they rank predictors according to
importance for classification accuracy. In the second scenario we
assess the agreement among rankings of the same patients according to
the predicted probabilities of having a malignant tumor. Here we apply
the same statistical classification method to XX bootstrap samples of
the MALOVA data to obtain XX prediction models. From each model we
predict the risks of a malignant patients in the DACOVA study.
Finally we estimate the agreement of the resulting XX rankings of the
DACOVA patients to illustrate the sensitivity of the corresponding
statistical classification method to pertubations of the learning
data. In the latter case we further investigate the behaviour of
different statistical methods in a simulation study.

@@

@@

While the sequential rank agreement is primarily an exploratory tool
we have suggested two null hypotheses that can be used to evaluate the
sequential rank agreement obtained. Note that none of the two null
hypotheses are concerned with the actual ``true ranking'' but are
purely concerned with consistency/stability of the rankings among the
lists. As such we cannot determine if the rankings are good but only
whether they agree. The sequential rank agreemenet curve can be
compared visually to the curves obtained under either of the null
distributions and simple point-wide $p$-values can be obtained for
each depth by counting the number of sequential rank agreements under
the null hypothesis that is less than or equal to the observed rank
agreement. Two simple extensions can be pursued to make more formal
uniform tests: one would be to use a Kolmogorov-Smirnov-like test and
use the largest difference between the variance-weighted sequential
rank agreement curve and the mean null rank agreement curve as a test
statistic. Alternatively, a change point analysis could be made on the
sequential rank agreement in order to determine the depths at which
there are ``jumps'' in the rank agreement. These jumps would
correspond to depths for which the agreement among the lists was
substantially worse and could serve as indicators for when the lists
no longer agree sufficiently satisfactory.

Finally, we have --- whenever possible --- used as much information
from the lists as possible, i.e., we have used all available ranks. If
we have information on the rankings as well as the corresponding
evidence for significance that we could choose to censor the rankings
of all the items that were deemed not to be significant. That would
ensure that there would be put less emphasis on the agreement of the
non-significant items and it would be easier to identify a change in
agreement among the items that were deemed to relevant.

@@

In conclusion we have introduced a method for evaluation of ranked
(censored) lists that can be easily interpreted and that can be
applied to a large number of situations.  The method can be adapted
even further by using it to identify and cluster statistical analysis
methods that provide more or less the same agreement, or by using the
rank agreement as a hyper-parameter in, say, elastic net regularized
regression where the rank agreement is used to determine the mixing
proportion between the $L_1$ and the $L_2$ penalty. Finally, the
proposed method may be adapted (with some additional assumptions) to
the situation where the lists are subsets of each other. We will be
investigating these extensions further in the future.

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[]{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[]{color}
\usepackage{subcaption}
\usepackage{natbib}
\usepackage{xspace}
\usepackage{calc}
\usepackage[noae]{Sweave}
\usepackage{authblk}
\usepackage{algorithm}
\usepackage{algpseudocode}
%\usepackage{geometry}
%\geometry{a4paper, portrait, margin=1in}
%
% New commands
%
\newcommand{\ie}{\emph{i.e.}\@\xspace}
\newcommand{\eg}{\emph{e.g.}\@\xspace}

@@

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and prominent task in applied statistics. Predictor
  variables are ranked according to magnitude of association with an
  outcome, prediction models rank subjects according to the
  personalized risk of an event, and genetic studies rank genes
  according to their difference in gene expression levels. This
  article constructs measures of the agreement of two or more ordered
  lists. We use the standard deviation of the ranks to define a
  measure of agreement that both provides an intuitive interpretation
  and can be applied to any number of lists even if some or all are
  incomplete or censored. The main idea is then a permutation based
  and graphically approach to assess the sequential changes of
  agreement as a function of the depth of the lists. This may help to
  indentify change-points in the agreement of the lists. The
  usefulness of these tools are illustrated using gene rankings, and
  using data from two Danish ovarian cancer studies on where we assess
  the within and between agreement of different statistical classification
  methods.
\end{abstract}

Key words: sequential rank agreement, partial ranking, order statistic, gene rankings,
methods comparison, variable selection, permutation

@@

Several approaches exist to measure the ``distance'' between two
ranked lists. Of these, Kendall's $\tau$ \citep{kendall1948} and
Spearman’s footrule $\rho$ \citep{Spearman1910} are among the most
well-known, but these measures do not distinguish between agreement in
the top versus towards the bottom of the lists. \citet{Shieh1998}
proposed a weighted version of Kendall's $\tau$ where each pair of
rankings can be assigned different weights, and \citet{Yilmaz2008}
proposed the $\tau_{ap}$ which places higher emphasis on the top of
the lists.
%
Spearman’s footrule uses the ranks of the variables for calculation of
the distance and the use of ranks are also employed in the $M$ measure
of \citet{Bar-Ilan2006} where the reciprocal rank differences are used
to calculate the agreement measure.

@@

In this article we introduce sequential rank agreement for measuring
agreement among ranked lists. We provide graphical tools to analyse
the depth of agreement of a set of ranked lists. The general idea is
to define agreement based on the sequence of ranks from the first $d$
elements in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{Carstensen2010}. Our proposed approach allows to compare more
than two lists simultaneously, provides a dynamic measure of agreement
as a function of the depth in the lists, places high weight on the top
of the list, accommodates censored/incomplete lists of varying
lengths, and has a natural interpretation that directly relates to the
ranks. The graphical illustration of sequential rank agreement
potentially allows to infer a change-point, i.e., a list depth where
there is a change in the agreement of the lists.

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$ and a
ranking function $R: \{X_1,\dots,X_P\}\to \{1,\dots,P\}$, such that
$R(X_p)$ is the rank of item $X_p$. The inverse mapping $R^{-1}$
assigns to rank $r\in\{1,\dots,P\}$ the item $R^{-1}(r)$ that was
given rank $r$. An ordered list is the realization of a ranking
function $R$ applied to a data sample $X$ and we let
$R_1(X),\dots,R_L(X)$, $L\geq2$, denote ordered lists obtained from
$L$ different ranking functions applied to the same data $X$. We will
use the same notation for the equivalent scenario where we have
ordered lists obtained from applying the same ranking function to $L$
samples from the same population, i.e., for ranked lists
$R(X^1), \ldots, R(X^L)$.  Panels (a) and (b) of
Table~\ref{tab:example} show a schematic example of these
mappings. Thus if $R^{-1}(1)=X_{34}$ then item $X_{34}$ is ranked
first in list $l$, and similarly $R_l(X_{34})=1$.

\begin{table}[tb]
  \caption{Example set of ranked lists. (a) shows the ranked list of
    items for each of three lists, (b) presents the ranks obtained by
    each item in each of the three lists and (c) shows the cumulative
    set of items up to a given depth in the three lists.}
\begin{center}
\footnotesize{
  \begin{subtable}{4.5cm}%
    \caption{}
      \begin{tabular}{cccc}
        \hline\hline % & \multicolumn{3}{c}{List} \\
        Rank & $R^{-1}_1$ & $R^{-1}_2$ & $R^{-1}_3$ \\ \hline
        1 & A & A & B \\
        2 & B & C & A \\
        3 & C & D & E \\
        4 & D & B & C \\
        5 & E & E & D \\ \hline
    \end{tabular}
  \end{subtable}
% \hfill
\hspace{1em}
%%%
  \begin{subtable}{4cm}%
    \caption{}
    \begin{tabular}{cccc} \hline\hline
  % & \multicolumn{3}{c}{List} \\
    Item & $R_1$ & $R_2$ & $R_3$ \\ \hline
    A & 1 & 1 & 2 \\
    B & 2 & 4 & 1 \\
    C & 3 & 2 & 4 \\
    D & 4 & 3 & 5 \\
    E & 5 & 5 & 3 \\ \hline
  \end{tabular}
\end{subtable}
% \hfill
\hspace{1em}
%%%
\begin{subtable}{4cm}%
  \caption{}
\begin{tabular}{cc} \hline\hline
% Depth   &     \\
Depth  &  $S_d$ \\ \hline
1 & $\{$A, B$\}$\\
2 & $\{$A, B, C$\}$ \\
3 & $\{$A, B, C, D, E$\}$ \\
4 & $\{$A, B, C, D, E$\}$ \\
5 & $\{$A, B, C, D, E$\}$ \\ \hline
\end{tabular}
\end{subtable}
}
\end{center}
\label{tab:example}
\end{table}

The agreement of the lists regarding the rank given to an item $X_p$ can be
measured by
\begin{equation}
  A(X_p) = f(R_1(X_p), \ldots, R_L(X_p)),
\end{equation}
for a distance function $f.$ Throughout this paper we will use the
sample standard error as our function $f$ and hence use
$$A(X_p) = \sqrt{\frac{\sum_{i=1}^L (R_i(X_p) - \bar{R}(X_p))^2}{L-1}},
$$
where $\bar{R}(X_p)$ is the average rank assigned to item $X_p$, but
other choices could be made. The sample standard error has an
interpretation as the average distance of the individual rankings of
the lists from the average ranking.

We now describe what is exemplified in Panel (c) of Table
\ref{tab:example} and how it can be used to define \emph{sequential
  rank agreement}. For an integer $1\le d\le P$ we define the unique
set of items found by merging the first $d$ elements of each of the
$L$ lists, i.e., the set of items ranked less than or equal to $d$ in
any of the lists:
\begin{equation}
S_d = \{R_l^{-1}(r) ; r \leq d, l = 1, \ldots, L \}.
\end{equation}
The \emph{sequential rank agreement} is the pooled standard deviation
of the items found in the set $S_d$:
\begin{equation}
\textrm{SRA}(d)= \sqrt{\frac{\sum_{\{p \in
      S_d\}}(L-1)A(X_p)^2}{(L-1)|S_d|}}, \label{def:sra}
\end{equation}
where $|S_d|$ refers to the cardinaly of the set $S_d$. Values of
\textrm{SRA} close to zero suggests that the lists agree while larger
values suggests disagreement. If the ranked lists are identical then
the value of SRA will be zero for all depths $d$. The sequential rank
agreement can be interpreted as the average distance of the individual
rankings of the lists from the average ranking for each of the items
we have seen until depth $d$. Note that the terms that are part of
\eqref{def:sra} are not independent since each list contains the ranks
from 1 to $P$ exactly once. Hence we simply use the pooled standard
deviation as a measure of the average distance between rankings.

@@

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to classify between acute lymphoblastic leukemia
(ALL) and acute myeloid leukemia (AML)
\citep{Golub1999}. Preprocessing of the gene expression data was done
as described in \citet{Dudoit2002} and the four different analysis
approaces were: marginal two-sample $t$ tests, marginal logistic
regression analyses, logistic regression eleastic net
\citep{friedman2010regularization}, and marginal maximum information
content correlations (MIC) \citep{Reshef2011}. For the first two
methods, the genes were ranked according to $p$ value, for logistic
regression the genes were ordered by size of the corresponding
coefficients (after standardization), and MIC was ordered by
correlation which resulted in the top rankings seen in Table~\ref{tab1}.

The sequential rank agreement plotted in Figure~\ref{fig:example1}
shows how many ranks apart we on average expect the genes found in the
top $d$ part of the lists to be for each depth $d$.  It can be seen
that agreement is higher in the top of the lists (smaller values on
the $y$ axis) corresponding to \emph{better} agreement than in the
tails of the lists.

@@

If the lists agree on all items then the sequential rank agreement is
 zero for all depths. Generally,  if there are changes in the level of rank
 agreement then this suggests that there are sets of items that are ranked
 similarly in all lists and other sets of items that have been assigned vastly
 different ranks in the lists. In gene association studies we
 generally expect the agreement among the lists to be better towards
 the top of the lists and worse towards the end of the lists. In this
 case the sequential rank agreement curve will start at a low level
 and then increase until it levels off as in
 Figure~\ref{fig:example1}.

@@

Incomplete lists are also a common occurrence. They arise for example
in case of missing data (items), when comparing top $d$ list results
from publications, or when some methods only rank a subset of the
items. For example, penalized regression based on the Lasso provides a
sparse set of predictors that have non-zero coefficients. There is no
obvious ordering of the set of predictors whose coefficient has been
shrunk to zero. Incomplete lists also occur when the analyst decides
to censor all lists at depth $d$ such that the measurement of
agreement is restricted to the first $d$ elements of each list.

Sequential rank agreement can be generalized to incomplete/censored
lists in the following way. Let $\Lambda_P$ be a set of $P$ items and
$\Lambda_l\subset \Lambda_P$ the subset of $d_l$ items that occur in
list $l$.  The case where all lists are censored at depth $d$
corresponds to $d_1=\cdots=d_L=d$. For incomplete/censored lists the
rank function becomes
\begin{equation}
  \tilde R_l(X_p) = \left\{\begin{array}{cl} \{R_l(X_p)\} & \text{ for } X_p\in \Lambda_l \\
      \{d_l+1,\dots,P\} & \text{ for } X_p \not\in \Lambda_l\end{array}\right.
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

@@

Like for fully observed lists we generally expect the sequential rank
agreement to start low and then increase unless the lists are
completely unrelated (in which case the sequential rank agreement will
be constant at a high level) or if the lists mostly agree on the
ranking (in which case the sequential rank agreement will also be
constant but at a low level). For censored lists we also expect a
change-point around the depth where the lists are censored. This is an
artefact stemming from the fact that we assume that the remainder of
the lists can be replaced by a simple permutation of the missing
items.

@@

The first reference hypothesis is
\begin{eqnarray*}
H_0  & : &  \text{The list rankings correspond to complete randomly}\\
       &  & \text{permuted lists}.
\end{eqnarray*}
This does not only assume that there is no information in data on
which the rankings are based but also that the methods are completely
independent.

@@

The overlap among $L$ lists observed until depth $d$ is defined as the
proportion of items that are found in all $L$ lists compared to the
possible number of items found in all lists observed until depth $d$:
\begin{equation}
  O(d) = \frac{| \cap_{l=1}^L \{R_l^{-1}(r); r\leq d \} |}{d}. \label{def:overlap}
\end{equation}
Typically the overlap is only considered for pairwise comparisons but
formula \eqref{def:overlap} can be applied even when $L>2$
\citep{Bar-Ilan2006,Boulesteix2009}.

@@

The MALOVA study is a multidisciplinary Danish study on ovarian cancer
\citep{Hogdall:2004:Cancer:15160342} where all Danish women diagnosed
with an ovarian tumor and referred for surgery from the participating
departments of gynecology were enrolled continuously from December
1994 to May 1999. For the purpose of illustration we use a random
subsample of $119$ patients with a total of $58$ patients with
malignant ovarian cancers as cases and $61$ patients with benign
ovarian tumors as controls. The DACOVA study is another
multidisciplinary Danish study on ovarian cancer which included about
$66\%$ of the female population of Denmark
\citep{bertelsen1991protocol}. It was aimed to continuously enroll all
patients that were referred to surgery of an ovarian tumor clinically
suspected to be cancer during the period 1984 to 1990.
% @Andreas: hvorfor random subsample?
Similarly, we use a random subsample of $113$
patients with a total of $54$ malignant ovarian cancers and $59$
benign ovarian tumors/gynecologic disorders.

Proteomic analysis was performed by MALDI-TOF mass spectrometry with a
Bruker Ultraflex in positive, linear ion mode in the mass-to-charge
range $800$ to $20000$ Dalton by an average of $500$ laser
shots. Calibration of the instrument was performed using a protein
standard mixture. Each spectrum consists of $49642$ samples over the
mass-to-charge range which we down-sample on an equidistant grid of
5000 points by linear interpolation. We further preprocess each
spectrum by first removing the slow-varying baseline intensities with
the SNIP algorithm \citep{ryan1988snip} followed by a normalization
with respect to the total ion counts. Finally, we standardize the 5000
predictors to have column-wise zero mean and unit variance.

We use these two data sets to illustrate how the sequential rank
agreement can be applied in two different scenarios. In the first
scenario we assess the agreement of different statistical
classification methods in how they rank predictors by different
statistical classification methods for the two study populations. In
the second scenario we assess the agreement among ranking of the
predicted individual risk probabilities of having malignant tumor
using four different prediction methods.
For this second example we furthermore
investigate the behaviour of different statistical methods in a
simulation study.

@@

Here we apply the same statistical classification
method to XX bootstrap samples of the MALOVA data to obtain four
prediction models. From each model we predict the risks of a malignant
patients in the DACOVA study.  Finally we estimate the agreement of
the resulting XX rankings of the DACOVA patients to illustrate the
sensitivity of the corresponding statistical classification method to
pertubations of the learning data.

@@

While the sequential rank agreement is primarily an exploratory tool
we have suggested two null hypotheses that can be used to evaluate the
sequential rank agreement obtained. Note that none of the two null
hypotheses are concerned with the actual ``true ranking'' but are
purely concerned with consistency/stability of the rankings among the
lists. As such we cannot determine if the rankings are good but only
whether they agree. The sequential rank agreemenet curve can be
compared visually to the curves obtained under either of the null
distributions and simple point-wise $p$-values can be obtained for
each depth by counting the number of sequential rank agreements under
the null hypothesis that is less than or equal to the observed rank
agreement. Two simple extensions can be pursued to make more formal
uniform tests: one would be to use a Kolmogorov-Smirnov-like test and
use the largest difference between the variance-weighted sequential
rank agreement curve and the mean null rank agreement curve as a test
statistic. Alternatively, a change-point analysis could be made on the
sequential rank agreement in order to determine the depths at which
there are ``jumps'' in the rank agreement. These jumps would
correspond to depths for which the agreement among the lists was
substantially worse and could serve as indicators for when the lists
no longer agree sufficiently satisfactory.

Finally, we have --- whenever possible --- used as much information
from the lists as possible, i.e., we have used all available ranks.
We could choose to censor the rankings of all the items that were
deemed not to be significant if information on both rankings and the
corresponding evidence for significance was available. That would
ensure that there would be put less emphasis on the agreement of the
non-significant items and it would be easier to identify a change in
agreement among the items that were deemed to relevant.

@@

In conclusion we have introduced a method for evaluation of ranked
(censored) lists that can be easily interpreted and that can be
applied to a large number of situations.  The method can be adapted
even further by using it to identify and cluster statistical analysis
methods that provide more or less the same agreement, or by using the
rank agreement to optimize a hyper-parameter in, say, elastic net regularized
regression where the rank agreement is used to determine the mixing
proportion between the $L_1$ and the $L_2$ penalty. Finally, the
proposed method may be adapted (with some additional assumptions) to
the situation where the lists are subsets of each other. We will be
investigating these extensions further in the future.
